---
title: "Blogdown draft"
author: "Nathanael Aff"
date: 2017-06-14T21:13:14-05:00
categories: ["How-to"]
tags: ["blogdown", "R"]
---
 
```{r global_options, echo = FALSE}
knitr::opts_chunk$set(
       fig.align = 'center', 
       echo=TRUE, 
       warning=FALSE, 
       message=FALSE, 
       results = "hide",
       fig.width = 6.5, 
       fig.height = 6.5,
       dev = 'png')
```
Requirements. 

Familiarity with R, R-studio, markdown, linux or Mac, or a 



# Blogdown 

[blogdown](https://github.com/rstudio/blogdown)

Basics : 
new_site() 
... etc. 

1. Create a directory where your website will live. 
2. Install the blogdown package. 
3. Create the default site. 


```
install.packages("blogdown") 
setwd("~/myblog")
blogdown::new_site()
# Check out the directory
dir()
```
### Install Theme 

The `new_site` function by default installs the hugo-lithium-theme into the themes folder. You can browse more themes [here](https://themes.gohugo.io/). I will be using the Minos theme. It's simple and supports Latex markup. 

The instructions for installing []()

```
git ...
git ... 
```

By default, this leaves a copy of the original git repository, so we need to remove that as well. 

```
rm -rf hugo-theme-minos/.git
```

### Config 

The default `config.toml` file should look like 

(FIXME) -- update after installation of theme
```
baseURL = "http://example.org/"
languageCode = "en-us"
title = "My New Hugo Site"

```

We've added a l

```
baseURL = "http://imagoodperson.com"
languageCode = "en-us"
title = "I am a good person"
ignoreFiles = ["\\.Rmd$", "_files$", "_cache$"]

[[menu.main]]
    name = "GitHub"
    url = "https://github.com/nateaff"

[params]
    katex = true
```

### Modify file 

```{r }


```






# Netlify

1. Authorize
2. Set up Domain 


# Domain 


## Modify fonts

https://fonts.google.com/


https://stoned.io/web-development/hugo/How-To-Dynamically-Use-Google-Fonts-In-A-Hugo-Website/

1. Add to head.html 

```
<link href='//fonts.googleapis.com/css?family={{ .Site.Params.fontbody }}' rel='stylesheet' type='text/css'>
<link href='//fonts.googleapis.com/css?family={{ .Site.Params.fontheading }}' rel='stylesheet' type='text/css'> 
```

At end of `head.html` add 
```
<style>
body {
    font-family: '{{ replace .Site.Params.fontbody "+" " "}}';
}

h1, h2, h3,
h4, h5, h6 {
    font-family: '{{ replace .Site.Params.fontheading "+" " "}}';
}
</style>
```

Here all heading sizes, h1,h2, etc., are being determiend by a single font. You can also repeat the aboves steps but specify the font for individual heading sizes.

Now we can specify the name of the font in the config in the `config.toml`. 

```
fontbody = "Tinos"
fontheading = "Lato"
```


















